---
title: "ENVS 193DS - Spring Final"
author: "Sascha Roberts"
date: "2026-06-09"
format: pdf
---

# Github Repo

My repo [here](https://github.com/Sascharr16/ENVS-193DS_spring-2026_final/tree/main)

# Set Up

## Load packages

```{r}
#| label: load_packages
#| message: false

# load in packages
library (tidyverse)
library (here)
library (janitor)
library (rstatix)
library (ggeffects)

```

## Load data

```{r}
#| label: load_data
#| message: false

nests <- read.csv (here("data", "nest_data_final.csv"))
# create object for bike commute data
bike_commute <- read.csv(here("data", "ENVS_193DS_Personal_Data.csv"))

```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, my coworker used _______. In part 2, they used _______.

## b. Suggestions for rewriting



## c. Further analysis



# Problem 2. Data analysis

## a. Clean and wrangle

```{r}
#| label: clean_nest_data

nests_clean <- nests |> 
  # select columns of interest 
  select (height_cm, case_control, ant_species) |> 
  # filter to only include ant species C. mimosae, C. sjostedti, and C. nigriceps
  filter (ant_species %in% c ("AB", "RRB", "BBR")) |> 
  # relevel ant species to be in the following order: 1) C. sjostedti, 2) C. mimosae, 3) C. nigriceps
  mutate (ant_species = fct_relevel (ant_species, "AB", "RRB", "BBR")) |> 
  # create a new column to store the full scientific name for each ant species
  mutate (scientific_name = case_when (
      ant_species == "AB"  ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"))

slice_sample(nests_clean, n = 10)
```



## b. Response variable



## c. Purpose of study



## d. Table of models



## e. Run the models



## f. Select the best model



## g. Check the diagnostics



## h. Visualize the model predictions



## i. Write a caption for your figure.



## j. Calculate model predictions



## k. Interpret your results



# Problem 3. Affective and exploratory visualizations

## a. Comparing visualizations

- My affective visualization does not include each individual observation as a data point like my exploratory data visualization does from homework 2. Instead, my affective visualization uses a bike path visual that represent the trend of the data and includes bikers overlaid to represent clusters of data points along the line.
- Both visualizations exhibit a clear positive linear shape that can be used to infer the relationship I discovered between bike commute duration and commute urgency. 
- Between both visualizations, there is a clear relationship between bike commute duration and commute urgency. The less time I have before class when I leave home, the quicker I bike from home to campus. 
- One piece of helpful feedback I received in workshop was that the way I represented outliers with the same object as my trendline was distracting and confusing, and I used this suggestion to create a more cohesive visualization by changing this representation. I also was recommended to better differentiate the bikers on the bottom left of my trendline to better represent that they are moving faster than other "points". I utilized this suggestion by adding dust clouds behind the "faster" bikers. 


## b. Designing an analysis

One statistical analysis that I could use to answer my research question is a linear model (linear regression). Because my central question of my data collection is; "How does the amount of time I leave my house before class effect the duration of my bike commute?", I can use a linear model to display this predicative relationship. My predictor variable is how much time I have from when I leave home until class start (minutes), and my response variable is the duration of bike commute (minutes). Both of these variables are continuous. I would expect these two variable to be connected and observe that my urgency to get to class predicts how quickly I ride my bike.


## c. Check any assumptions and run your analysis

### Clean data

```{r}
#| label: clean_bike_data

# create new object for clean data
bike_commute_clean <- bike_commute |> 
  # clean column names
  clean_names() |> 
  # select columns of interest 
  select(duration_m,
         time_before_class_min)

```


### Visualize data

```{r}
#| label: personal_data_visualization

# base layer: ggplot
# data frame: bike_commute_clean
ggplot (data = bike_commute_clean,
      mapping = aes(
      # x-axis: amount of time before class
      # y-axis: duration of commute
      x = time_before_class_min,
      y = duration_m)) +
  # first layer: point = scatterplot
  # make data points colored, change shape
  geom_point(color = "blue3",
             shape = 21) +
  # Changed title, x and y-axis
  labs (title = "Urgency of Classtime Bike Commute",
        x = "Amount of Time Before Class",
        y = "Duration of Commute (minutes)") +
  # applied custom plot theme
  theme_bw() +
  # removing the panel background
  theme(panel.background = element_blank()) +
  # removing the panel grid
  theme(panel.grid = element_blank()) +
  # removing axis ticks
  theme(axis.ticks = element_blank())
```


### Independent observations

My data has independent observations. Each individual observation is independent of the observations before that. My bike commute duration to class one day, does not effect my next commute duration.

### Write out model

```{r}
#| label: fit_bike_model

# create new object for bike commute
bike_commute_model <- lm (duration_m ~ time_before_class_min, # duration as a function of time before class
                          data = bike_commute_clean) # data: bike_commute_clean

```

### Check assumptions with diagnostic plots

```{r}
#| label: diagnostic_plots

# display diagnostic plots in 2 by 2 grid
par (mfrow = c(2, 2))
# plot bike commute model
plot (bike_commute_model)


```

**Note:** Using the "Residuals vs Fitted" and "Scale-Location" plots, we can determine that the residuals are evenly distributed and have constant variance (homoscedasticity). Using the "QQ Residuals" plot we can determine that the residuals are normally distributed. Finally, using the "Residual vs Leverage" plot, we can determine that there are no outliers that are influencing our model. 

**All assumptions are met**

### Model summary

```{r}
#| label: commute_model_summary

#bike commute model summary
summary (bike_commute_model)

```

**Note:**

Intercept: $0.36905 \pm 0.4897$

Slope: $0.50063 \pm 0.0473$

Adjusted R-squared: 0.8538

P-value: p < 0.0001

## d. Create a visualization

### Model predictions

```{r}
#| label: commute_predictions
#| message: false

# new object call bike_commute_preds
bike_commute_preds <- ggpredict (bike_commute_model, # model object
                                 terms = "time_before_class_min") # predictor

# display the predictions
bike_commute_preds
```

### Visualization

```{r}
#| label: bike_commute_final_visualization

# base layer: ggplot
# using clean data frame
ggplot(data = bike_commute_clean,
       aes(x = time_before_class_min,
           y = duration_m)) +
  # first layer: points representing individual observations
  geom_point(size = 2,
             stroke = 0.5,
             fill = "cyan4",
             shape = 21) +
  # second layer: ribbon representing confidence interval
  # using predictions data frame
  geom_ribbon(data = bike_commute_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.1) +
  # third layer: line representing model predictions
  # using predictions data frame
  geom_line(data = bike_commute_preds,
            aes(x = x,
                y = predicted)) +
  # axis labels
  labs( title = "Higher urgency predicts shorter bike commute duration",
       x = "Amount of Time Before Class Starts (minutes)", 
       y = "Duration of Commute (minutes)") +
  # theme
  theme_minimal()
```


## e. Write a caption

**Figure 2: Leaving home with less time before class predicts a shorter bike commute** Visualization displays linear relationship between the "Amount of Time Before Class Starts (minutes)" on the x-axis, and the "Duration of Commute (minutes)" on the y-axis. Blue points represent individual bike commute observations. Line represents model predictions, and ribbon displays 95% confidence interval.

## f. Write about your results

My biking commute duration increased as the amount of time I had before class increased (linear regression, F(1, 18) = 112, p < 0.0001, $\alpha$ = 0.05). 85% ($R^2$ = 0.85) of the variation in the model can be explained the predictor.

Based on model predictions, for every 1 minute increase in available time before class starts, I expect an increase in commute duration of $0.5 \pm 0.047$ (SE) minutes.

## g. Sharing your affective visualization

I attended week 10 workshop! :)

