---
title: "ENVS 193DS - Spring Final"
author: "Sascha Roberts"
date: "2026-06-09"
format: pdf
---

# Github Repo

My repo [here](https://github.com/Sascharr16/ENVS-193DS_spring-2026_final/tree/main)

# Set Up

## Load packages

```{r}
#| label: load_packages
#| message: false

library (tidyverse)
library (here)
library (janitor)
library (rstatix)

```

## Load data

```{r}
#| label: load_data
#| message: false

bike_commute <- read.csv(here("data","ENVS_193DS_Personal_Data.csv"))

```

# Problem 1. Research writing

## a. Transparent statistical methods

In part 1, my coworker used _______. In part 2, they used _______.

## b. Suggestions for rewriting



## c. Further analysis



# Problem 2. Data analysis

## a. Clean and wrangle



## b. Response variable



## c. Purpose of study



## d. Table of models



## e. Run the models



## f. Select the best model



## g. Check the diagnostics



## h. Visualize the model predictions



## i. Write a caption for your figure.



## j. Calculate model predictions



## k. Interpret your results



# Problem 3. Affective and exploratory visualizations

## a. Comparing visualizations

- My affective visualization does not include each individual observation as a data point like my exploratory data visualization does from homework 2. Instead, my affective visualization uses a bike path visual that represent the trend of the data and includes bikers overlaid to represent clusters of data points along the line.
- Both visualizations exhibit a clear positive linear shape that can be used to infer the relationship I discovered between bike commute duration and commute urgency. 
- Between both visualizations, there is a clear relationship between bike commute duration and commute urgency. The less time I have before class when I leave home, the quicker I bike from home to campus. 
- One piece of helpful feedback I received in workshop was that the way I represented outliers with the same object as my trendline was distracting and confusing, and I used this suggestion to create a more cohesive visualization by changing this representation. I also was recommended to better differentiate the bikers on the bottom left of my trendline to better represent that they are moving faster than other "points". I utilized this suggestion by adding dust clouds behind the "faster" bikers. 


## b. Designing an analysis

One statistical analysis that I could use to answer my research question is a linear model (linear regression). Because my central question of my data collection is; "How does the amount of time I leave my house before class effect the duration of my bike commute?", I can use a linear model to display this predicative relationship. My predictor variable is how much time I have from when I leave home until class start (minutes), and my response variable is the duration of bike commute (minutes). Both of these variables are continuous. I would expect these two variable to be connected and observe that my urgency to get to class predicts how quickly I ride my bike.


## c. Check any assumptions and run your analysis

### Visualize data

```{r}
#| label: personal_data_visualization

# base layer: ggplot
# data frame: bike_commute
ggplot (data = bike_commute,
mapping = aes(
# x-axis: amount of time before class
# y-axis: duration of commute
x =
`time_before_class_(min)`
,
`duration_(m)`
y =
)) +
8
# first layer: point = scatterplot
# make data points colored, change shape
geom_point(color = "blue3",
shape = 21) +
# add reference linear line that fits model, make line a lighter blue
geom_smooth(method = "lm", se = FALSE,
color = "deepskyblue") +
# Changed title, x and y-axis
labs (title = "Urgency of Classtime Bike Commute",
x = "Amount of Time Before Class",
y = "Duration of Commute (minutes)",
subtitle = "2026-05-28") +
# applied custom plot theme
theme_bw() +
# removing the panel background
theme(panel.background = element_blank()) +
# removing the panel grid
theme(panel.grid = element_blank()) +
# removing axis ticks
theme(axis.ticks = element_blank())
```


### Indpendent observations

My data has indepredent observations. Each individual observation is independent of the obseravtions beofre that. My bike commute duration to class one day, does not effect my next commute duration.

#### 


## d. Create a visualization



## e. Write a caption



## f. Write about your results



## Sharing your affective visualization

